From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments

### 2015-09-14 write_up.tex
edit before olga edit

High-throughput omics technologies enable large-scale experiments, with
simultaneous measurements of multiple cellular components in a sample.
The huge amount of data generated paradoxically make the task of causal
inference much more difficult, because, when many associations are
examined, some will appear significant just by chance. In other words,
large-scale experiments are more prone to hiding the true signal and
generating spurious associations, leading to increased false positive
conclusions that causal relationships are present. This problem can be
addressed by improving experimental design and/or refining the
biological question. Below, we describe the process of elucidating
statistical associations from data and suggest practical approaches for
analysis of large scale datasets.
\section{Discussion}
\subsection{I. Statistical inference of causal relationships}

In MAPK signaling, Raf causally affects active Mek levels (via
phosphorylation), while Mek causally affects Erk. Imagine this
relationship was unknown: could it be detected from measurements of
these phosphoproteins? Given sufficient data, a correlation will be
detected between each pair of proteins, due to (direct and indirect)
causal relationships. Figure 1, top panel, depicts these correlation as
black lines, causal relationships are depicted as blue arrows.

@@

First, some terminology. A statistical association (such as a
correlation) between two variables indicates that they are dependent.
Sometimes we choose to assess a dependence in the context of the value
of other system components, such as a third variable. In this case the
dependence is called conditional dependence, because it is evaluated
conditional on the state of other system components (i.e., the third
variable). Let's see how this applies here, by examining the dependence
between Raf and Erk. If we condition on Mek, then we assume that we know
the value of Mek: perhaps we select only samples which have a high level
of Mek, so Mek is fixed at level ``high''. Can the dependence between
Raf and Erk still be detected? If, due to noise, some of these Mek=high
samples have an intermediate or even low level of Raf, will Erk be
correspondingly low? Based on the causal mechanism, Erk will more
closely follow Mek, since Mek is its causal parent. Because of that, the
Raf-Erk dependence, once conditioned on Mek, will be reduced or
eliminated, thus indicating that the Raf-Erk edge can be removed. Using
this approach, it is possible to keep only the causal edges as depicted
by the black lines in Figure 1, bottom panel.

Computational methods that do causal inference in the context of an
experiment that simultaneously measures multiple intercorrelated
response variables all essentially have the same two step process; (1)
reduce the dense set of pairwise associations to a sparse set of
conditional dependencies, (2) use information about the interventions
included in the experiment to evaluate those conditional dependencies as
evidence for potential causal relations.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
\caption{As you increase the number of features, you get more high scoring
spurious correlations.}
\label{spur_corr}
\end{figure}

@@

As you increase the number of features you measure, the number of false
positive detection of conditional dependence explodes. What this means
is that the computational methods for causal inference will fail for the
typical large-scale experiment, because when their input data contains a
large amount of features relative to sample size, they cannot reliably
detect the conditional dependence relationships they need to infer
causality.

The second problem is the requirement for interventions. As the number
of features grows, the number of interventions needed to infer causality
grows, and performing sufficient perturbation experiments becomes
infeasible.

@@

connections a priori can help reduce the number of possible

>>>>

High-throughput omics technologies enable large-scale experiments, with
simultaneous measurements of multiple cellular components in a sample.
The huge amount of data generated paradoxically make the task of causal
inference much more difficult, because, when many associations are
examined, some will appear significant just by chance. In other words,
large-scale experiments are more prone to hiding the true signal and
generating spurious associations, leading to increased false positive
conclusions that causal relationships are present. This problem can be
addressed by improving experimental design and/or refining the
biological question. Below, we describe the process of elucidating
statistical associations from data and suggest practical approaches for
analysis of large scale datasets.

\section{Discussion}
\subsection{I. Statistical inference of causal relationships}

In MAPK signaling, Raf causally affects active Mek levels (via
phosphorylation), while Mek causally affects Erk. Imagine this
relationship was unknown: could it be detected from measurements of
these phosphoproteins? Given sufficient data, a correlation will be
detected between each pair of proteins, due to (direct and indirect)
causal relationships. Figure 1, top panel, depicts these correlations as
black lines, causal relationships are depicted as blue arrows.

@@

First, some terminology. A statistical association (such as a
correlation) between two variables indicates that they are dependent.
Sometimes we choose to assess a dependence in the context of the value
of other system components, such as a third variable. In this case the
dependence is called conditional dependence, because it is evaluated
conditional on the state of other system components (i.e., the third
variable). Let's see how this applies here, by examining the dependence
between Raf and Erk. If we condition on Mek, then we assume that we know
the value of Mek: perhaps we limit our analysis to samples which have a high level
of Mek, so Mek is fixed at level "high". Can the dependence between
Raf and Erk still be detected? If, due to noise, some of these Mek=high
samples have an intermediate or even low level of Raf, will Erk be
correspondingly low? Based on the causal mechanism, Erk will more
closely follow Mek (i.e. its values will tend to be high), since Mek is 
its causal parent. Because of that, the Raf-Erk dependence, once conditioned 
on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge 
can be removed. Using this approach, it is possible to keep only the causal edges 
as depicted by the black lines in Figure 1, bottom panel.

Computational methods for causal inference in the context of
experiments that simultaneously measure multiple intercorrelated
features all rely on finding conditional dependence.  All of these algorithms
essentially have the same two step process; (1)
reduce the dense set of pairwise associations to a sparse set of
conditional dependencies, (2) use information about the interventions
included in the experiment to evaluate those conditional dependencies as
evidence for potential causal relations.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
\caption{Increasing the number of features results in more high scoring
spurious correlations.}
\label{spur_corr}
\end{figure}

@@

As we increased the number of features we measured, we saw a comparable increase in 
the number of false positive detection of conditional dependence relationships. What this
means is that the computational methods for causal inference will fail for the
typical large-scale experiment, because when their input data contains a
large amount of features relative to sample size, they cannot reliably
detect the conditional dependence relationships they need to infer
causality.

The second problem is the requirement for interventions. This is just as important as the 
problem of spurious conditional dependence, but much easier to explain -- as the number
of features grows, the number of interventions needed to infer causality
grows, and performing sufficient perturbation experiments becomes
infeasible.

@@

connections are present in the data a priori can help reduce the number of possible
